Statistical methods for relative quantification of post-translational modifications in global proteomics experiments

### 2021-06-02 supplementary/ptm_sm.tex
some comments 06/02

With protein-level adjustment, the model-based testing is based on the hypothesis that there is no difference in adjusted PTM abundance between Conditions $i$ and $i'$
%Adjusted by protein abundance, the hypothesis states that there is no difference in log-abundance of the modification between groups $i$ and $i'$
\begin{align*}
H_{0}: \Delta = \left( \mu_{i} - \mu_{i'} \right) - \left( \mu_{i}^{\ast} - \mu_{i'}^{\ast} \right) &= 0 \\
H_{a}: \Delta = \left( \mu_{i} - \mu_{i'} \right) - \left( \mu_{i}^{\ast} - \mu_{i'}^{\ast} \right) &\neq 0
\end{align*}
The log-fold change in the adjusted PTM abundance, $\Delta$, is estimated by 
\[
\hat{\Delta} = \left[ \frac{1}{J} \left( \hat{y}_{i+} - \hat{y}_{i'+} \right) \right] - \left[ \frac{1}{J} \left( \hat{y}_{i+}^{\ast} - \hat{y}_{i'+}^{\ast} \right) \right],
\]
and the standard error of the estimate $\mathrm{SE}(\hat{\Delta})$ is 
\[
\left[ \frac{2}{J} \left( \hat{\sigma}_{\gamma}^{2} + \hat{\sigma}_{\gamma^{\ast}}^{2} \right) \right]^{1/2}.
\]
%\[
%\left[ \left( \frac{2}{J} \right) \cdot \left( \hat{\sigma}_{\gamma}^{2} + \hat{\sigma}_{\gamma^{\ast}}^{2} \right) \right]^{1/2}.
%\]
The test statistic $\hat{\Delta} / \mathrm{SE}(\hat{\Delta})$ is compared against the $t$ distribution, with degrees of freedom approximated by
\[
\left( \hat{\sigma}_{\gamma}^{2} + \hat{\sigma}_{\gamma^{\ast}}^{2} \right)^2 \bigg/
\left( \frac{\hat{\sigma}_{\gamma}^{4}}{\mathrm{df}(\gamma)} + \frac{\hat{\sigma}_{\gamma^{\ast}}^{4}}{ \mathrm{df}(\gamma^{\ast})} \right).
\]
A distinctive property of the proposed model-based testing to the two-sample $t$-test is that even only the PTM abundances in Conditions $i$ and $i'$ are compared, measurements from all conditions are used for the modeling and inference.

@@

\begin{figure}[h!]
\centering
\includegraphics[width=\textwidth]{sim/prot_pwr}
%\includegraphics[width=.85\textwidth]{sim/prot_pwr}
\caption{In comparison between the proposed approach and $t$-test with protein-level adjustment, the proposed approach improved power with small sample sizes. Two-sample $t$-test only used data within the groups of interest while ignoring the rest the of data. Consequently, it gave similar performance across cases with different number of conditions. In contrast, the proposed approach leveraged all available information, which resulted in improved power with increased number of conditions. \label{fig:prot_pwr}}
\end{figure}

@@

@@

The proposed approach adjusts for the underlying protein abundance in the PTM significance analysis, which corrects the confounding factor with a cost of increased uncertainty. We compared the required sample size with or without the adjustment to highlight the property (\sfigref{fig:size_prot}), based on a design of two conditions, in consideration of three pairs of standard deviations of log-intensities for modified and unmodified peptides: (0.2, 0.1), (0.2, 0.2), and (0.2, 0.4). 
%We considered a design with two conditions, where standard deviations of log-intensities of both modified and unmodified peptides were both set as $0.2$. 
We then characterized the advantage of general statistical modeling in complex designs in terms of sample size calculation (\sfigref{fig:size_grp}) and power analysis (\sfigref{fig:pwr_grp}).

>>>>

With protein-level adjustment, the model-based testing is based on the hypothesis that there is no difference in adjusted PTM abundance between Conditions $i$ and $i'$
%Adjusted by protein abundance, the hypothesis states that there is no difference in log-abundance of the modification between groups $i$ and $i'$
\begin{align*}
H_{0}: \Delta = \left( \mu_{i} - \mu_{i'} \right) - \left( \mu_{i}^{\ast} - \mu_{i'}^{\ast} \right) &= 0 \\
H_{a}: \Delta = \left( \mu_{i} - \mu_{i'} \right) - \left( \mu_{i}^{\ast} - \mu_{i'}^{\ast} \right) &\neq 0
\end{align*}
The log-fold change in the adjusted PTM abundance, $\Delta$, is estimated by 
\[
\hat{\Delta} = \left[ \frac{1}{J} \left( \hat{y}_{i+} - \hat{y}_{i'+} \right) \right] - \left[ \frac{1}{J} \left( \hat{y}_{i+}^{\ast} - \hat{y}_{i'+}^{\ast} \right) \right],
\]
and the standard error of the estimate $\mathrm{SE}(\hat{\Delta})$ is 
\[
SE(\hat{\Delta}) = \left[ \frac{2}{J} \left( \hat{\sigma}_{\gamma}^{2} + \hat{\sigma}_{\gamma^{\ast}}^{2} \right) \right]^{1/2}.
\]
%\[
%\left[ \left( \frac{2}{J} \right) \cdot \left( \hat{\sigma}_{\gamma}^{2} + \hat{\sigma}_{\gamma^{\ast}}^{2} \right) \right]^{1/2}.
%\]
The test statistic $\hat{\Delta} / \mathrm{SE}(\hat{\Delta})$ is compared against the $t$ distribution, with degrees of freedom approximated by
\[
\left( \hat{\sigma}_{\gamma}^{2} + \hat{\sigma}_{\gamma^{\ast}}^{2} \right)^2 \bigg/
\left( \frac{\hat{\sigma}_{\gamma}^{4}}{\mathrm{df}(\gamma)} + \frac{\hat{\sigma}_{\gamma^{\ast}}^{4}}{ \mathrm{df}(\gamma^{\ast})} \right).
\]
A distinctive property of the proposed model-based testing to the two-sample $t$-test is that even only the PTM abundances in Conditions $i$ and $i'$ are compared, measurements from all conditions are used for the modeling and inference.

@@

\begin{figure}[h!]
\centering
\includegraphics[width=\textwidth]{sim/prot_pwr}
%\includegraphics[width=.85\textwidth]{sim/prot_pwr}
\caption{In comparison between the proposed approach and $t$-test with protein-level adjustment, the proposed approach improved power with small sample sizes. Two-sample $t$-test only used data within the groups of interest while ignoring the rest the of data. Consequently, it gave similar performance across cases with different number of conditions. In contrast, the proposed approach leveraged all available information, which resulted in improved power with increased number of conditions.\todo{how about protein abundance change?} \todo{this conclusion seems that overfitting.. more number of conditions increase the power.. Then how about limma?} \todo{how to calculate estimated sigma for 4 conditions and t-test?} \label{fig:prot_pwr}}
\end{figure}

@@

\todo{should address number of replicates, standard deviations, missing data?, same as one site per protein.}
\todo{not strong evidence. need to revisit this section.}

@@

The proposed approach adjusts for the underlying protein abundance in the PTM significance analysis, which corrects the confounding factor with a cost of increased uncertainty. \todo{uncertainty? means variation?} We compared the required sample size with or without the adjustment to highlight the property (\sfigref{fig:size_prot}), based on a design of two conditions, in consideration of three pairs of standard deviations of log-intensities for modified and unmodified peptides: (0.2, 0.1), (0.2, 0.2), and (0.2, 0.4). 
%We considered a design with two conditions, where standard deviations of log-intensities of both modified and unmodified peptides were both set as $0.2$. 
We then characterized the advantage of general statistical modeling in complex designs in terms of sample size calculation (\sfigref{fig:size_grp}) and power analysis (\sfigref{fig:pwr_grp}).
